Fail Faster

### 2025-03-23 paper/paper.tex
through 3.5

\subsection{Design of a Staged Generator DSL}
\label{subsection:basic-design}
To build a staged version of a generator DSL, we must rewrite the the generator
combinators to do as much as possible at compile time.
The compile-time stage will then produce simplified code, free of any DSL combinators, which can be compiled and run
with different sizes and seeds.
Our job is thus to carefully bisect the DSL, determining which inputs to generator
combinators are known statically (and can be part of the compile time stage) and
which parts are only known at run time (and must be \texttt{code}).
In the staging literature, this task is known as ``binding-time analysis'' and it is more an art than a science.

@@

The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
known at compile time are the particular random seed and
size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
This leads us to define our library's generator type \texttt{'a
Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
function from dynamically-known size and seed to dynamically-determined result.

The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that runs \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

  let return (cx : 'a code) : 'a t = fun size_c random_c -> cx

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
allocate lists in the hot path of the generator. Even though these lists are usually small ---
at most a few dozen elements --- each allocation takes us closer to the next garbage collection.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s specialized to the list of weights known at compile time, that searches for the selected generator.

@@

Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
Given a list (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
traverses the tree of three \texttt{if}s to find the correct value to return.

@@

In Figure~\ref{fig:codecps-and-final-gen}, we follow prior work \cite{kovacs24, carrette05} and define
the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
In prior work, this type is often referred to as the ``code generation'' monad.
This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
\texttt{CodeCps.run} passes the continuation transformer the identity continuation to produce a \texttt{'a code}.
To avoid confusion with random data generators, we refer to this type as \texttt{CodeCps.t}.
Most importantly, the \texttt{CodeCps} type supports a function \texttt{let\_insert}, which, given \texttt{cx : 'a code},
let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the
continuation.
\hg{I'm getting lost in the inline code again}

We can then redefine\footnote{
The type \texttt{'a Gen.t} from before is like \texttt{'a code Gen.t} with this new definition. This type change
carries through all of our combinators. For example, \texttt{Gen.int} now returns \texttt{int code Gen.t}.
} our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.

This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
individual combinators must be carefully written to ensure that \texttt{'a code} values that contain effects are \texttt{let\_insert}'d.
To ensure that the library is written correctly, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
By differentially testing \cn{} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.

@@

  let int (lo : int code) (hi : int code) : int code t =
    fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
end
\end{lstlisting}
\caption{CodeCPS and The Final Gen Monad}
\label{fig:codecps-and-final-gen}
\end{figure}

>>>>

\subsection{Design of a Staged Generator DSL}
\label{subsection:basic-design}
To build a staged version of a generator DSL, we must rewrite the the generator
combinators to do as much as possible at compile time.
The compile-time stage will then produce simplified code, free of any DSL abstraction, that can be compiled and run
with different sizes and seeds.
Our job is thus to carefully bisect the DSL, determining which inputs to generator
combinators are known statically (and can be part of the compile time stage) and
which parts are only known at run time (and must be treated as \texttt{code}).
In the staging literature, this task is known as ``binding-time analysis,'' and it is more art than science.

@@

The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
known at compile time are the random seed and
size parameters. In practice, generators themselves are entirely known at compile time.
This leads us to define our library's generator type \texttt{'a
Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
function from dynamically known size and seed to dynamically determined result.

The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that runs \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size random}
and \texttt{k .<a>. size random} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' 
from Section~\ref{subsection:msp}.
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

  let return (cx : 'a code) : 'a t = fun size random -> cx

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
allocate lists in the hot path of the generator. Even though these lists are usually small ---
at most a few dozen elements --- each allocation takes us closer to the next garbage collection, which is bad for performance.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile-time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.

@@

Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
Given a list of (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and
traverses the tree of three \texttt{if}s to find the correct value to return.

@@

In Figure~\ref{fig:codecps-and-final-gen}, following prior work \cite{kovacs24, carrette05}, we define
the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}---a fact that deserves further investigation.}.
The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
In prior work, this type is often referred to as the ``code generation'' monad.
This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
\texttt{CodeCps.run} passes the continuation transformer the identity continuation to produce a \texttt{'a code}.
To avoid confusion with random data generators, we refer to this type as \texttt{CodeCps.t}.
Most importantly, the \texttt{CodeCps} type supports a function \texttt{let\_insert}, which, given \texttt{cx : 'a code},
let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the
continuation.
\hg{I'm getting lost in the inline code again}

We can then redefine
our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
The (old) type \texttt{'a Gen.t} is now written as the (new) type \texttt{'a code Gen.t}, and this type change
carries through all of our combinators. For example, \texttt{Gen.int} now returns \texttt{int code Gen.t}.

This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
For example, \texttt{bind (int .<0>. .<1>.) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}
This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
individual combinators must be carefully written to ensure that \texttt{'a code} values that contain effects are \texttt{let\_insert}ed.
To validate the library, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
By differentially testing \cn{} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.

@@

  let int (lo : int code) (hi : int code) : int code t =
    fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
end
\end{lstlisting}
\caption{CodeCPS and the Final Gen Monad}
\label{fig:codecps-and-final-gen}
\end{figure}
